---
title: "Untitled"
author: "Ghislaine van Gilse"
date: "10/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=T, eval=F}
# update conda to the newest version and answer y on question proceed
echo "y" | conda update -n base conda

```

```{bash, engine.opts='-i', echo=T, eval=T}
# show all conda virtual environments
conda env list

# show yml file content
cat ~/daur2/setup_meta_env.yml

```

```{bash, engine.opts='-i', echo=T, eval=F}
# create conda virtual environment
conda env create --file ~/daur2/setup_meta_env.yml
```

Exercise 6.1
Can you think of a reason that we specify the python dependency? Python is already installed, otherwise we wouldn’t be able to use the conda software, which is python based software…

  Software can be quite dependent on the particular version of python that is used.
  Later on we will need to use software that is dependent on this version of python, which is not the default version when opening a conda environment.
  

Exercise 6.2
The read length of this data is about 100-150 bp. You can see this using the following command:
zcat HU1_MOCK1_L001_R1_001.fastq.gz | head -n 10
Google the typical read length range for illumina and minion.
What sequencing platform was used to generate this dataset with?
  The illumina. Illumina has reads around 150bp while minion has way longer reads

Exercise 6.3
What could R1 and R2 stand for?
  Forward and reverse reads. Every forward read has a corresponding reverse read.

```{bash, engine.opts='-i', echo=T, eval=F}
# activate environment
conda activate meta

# install fastqc and answer yes for procceed question
echo "y" | conda install -c bioconda fastqc

# deactivate conda virtual environment
conda deactivate
```

```{bash, engine.opts='-i', echo=T, eval=F}
# activate environment
conda activate meta

# perform FastQC analysis (change first path to your own folder structure if needed)
fastqc -o ~/daur2/les6 /home/daur2/metagenomics/reader_data/HU1_MOCK1_L001_R1_001.fastq.gz
fastqc -o ~/daur2/les6 /home/daur2/metagenomics/reader_data/HU1_MOCK1_L001_R2_001.fastq.gz

# deactivate conda virtual environment
conda deactivate
```

Exercise 6.4
a) How many reads are in your dataset?
  40237621 reads.
b) What is the average quality?
  Zo rond de 36?
c) How many reads are of poor quality?
  0

Exercise 6.5
What is the average error rate in percentages?
  >99,9% accuracy


Exercise 6.6
Based on the quality analysis: do we need to filter out particular reads or do we need to trim particular bases for all reads?
  No
  
```{r, library}
library(png)
library(grid)
library(gridExtra)
```

```{r , echo=FALSE, fig.height=3, out.width= '100%', fig.cap="Fig. 1: Quality distribution: Right-side) Foward read quality scores range from 0 to 40 on the Phred scale. All bases for all reads have very high quality (>30), indicating that our reads have an accuracy of >99,9%. Left-side)" }
img1 <-  rasterGrob(as.raster(readPNG("~/daur2/les6/daur2_les6_qualityplot.png")))
img2 <-  rasterGrob(as.raster(readPNG("~/daur2/les6/daur2_les6_qualityplot2.png")))
grid.arrange(img1, img2, ncol=2, top=textGrob("Per base quality distribution of forward (right) and reverse (left) reads", gp=gpar(fontsize=10,font=8)))
```

Exercise 6.7
Do all bases have equal quality of the entire read?
  No, the final five bases on the reverse reads have decreased quality and a higher variability. However, the average quality (blue line) is still above Phred score 30, hence we do not need to remove those from our data set.
  
Exercise 6.8
How do we include creating our virtual environment and FastQC analysis inside the rmd file, without rerunning the code every time we knit our html report?
  To avoid rerunning analysis that are already finished you need to set eval=FALSE in the chunk header.