---
title: "daur2_les4"
author: "Ghislaine van Gilse"
date: "10/19/2022"
output: html_document
---

```{r setup, include=FALSE}

library(Rsubread)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(pheatmap)
library("org.Hs.eg.db")
library(GO.db)
library(GOstats)

```

```{r top10 genes}
# Create a dataframe for the 10 most significantly DE genes
top10_genes <- airway_dge_results[order(airway_dge_results$padj)[1:10],] %>% data.frame()
top10_genes
```
```{r rownames}
columns(org.Hs.eg.db)

help("SYMBOL")

# Create an extra column with the Entrez id (based on rownames)
top10_genes <- top10_genes %>% mutate(entrezid = rownames(top10_genes))

# Obtain the gene symbols based on the Entrez identifiers
top10_genes$symbol <- mapIds(org.Hs.eg.db,
                             keys = top10_genes$entrezid,
                             column = "SYMBOL",
                             keytype = "ENTREZID",
                             multiVals = "first")

head(top10_genes)
```

Exercise 4.1
Look at the help page of the mapIds() function. Can you explain how this function works and what the different options of the function do?
x	
the AnnotationDb object. But in practice this will mean an object derived from an AnnotationDb object such as a OrgDb or ChipDb object.

keys	
the keys to select records for from the database. All possible keys are returned by using the keys method.

columns	
the columns or kinds of things that can be retrieved from the database. As with keys, all possible columns are returned by using the columns method.

keytype	
the keytype that matches the keys used. For the select methods, this is used to indicate the kind of ID being used with the keys argument. For the keys method this is used to indicate which kind of keys are desired from keys


The function looks up for each gene identifier (the keys option can be used to specify the gene identifiers of interest, and the keytype option to specify the type of gene identifier) the annotation of interest (which is specified by the column option). In case of multiple matching annotations (in this case, one Entrez identifier matching more than one gene symbols), the multiVals = "first" option specifies that only the first match should be reported.


Exercise 4.2
One of the genes in the top 10 of most significantly DE genes is DUSP1. Read the abstract of this article. Based on this article, is it surprising to find DUSP1 in the top 10 or is it as expected? Explain your answer.
  The article shows results that DUSP1 is regulated by glucocorticoids. Dexamethasone is a glucocorticoid. It is therefore expected that DUSP1 is upregulated by the dexamethasone treatment. Indeed, we find a LFC of 2.9, indicating that DUSP1 is almost 8 times upregulated in response to the treatment.

Exercise 4.3
Use the mapIds() function to look up the cytogenetic bands for the 10 most significantly DE genes to figure out on which chromosomes the genes are located. Add this information as an extra column to the dataframe.
```{r chromosomal band}
top10_genes$chromosomal_band <- mapIds(org.Hs.eg.db,
                             keys = top10_genes$entrezid,
                             column = "MAP",
                             keytype = "ENTREZID",
                             multiVals = "first")

head(top10_genes)
```

```{r Obtaining GO terms}
top_upregulated <- top10_genes[which.max(top10_genes$log2FoldChange),"entrezid"]

# Find the associated GO terms
top_upregulated_GOterms <- select(org.Hs.eg.db,
                                  keys = top_upregulated,
                                  column = c("GO", "ONTOLOGY"),
                                  keytype = "ENTREZID",
                                  multiVals = "list")

# Select only the GO terms that are associated with biological processes
top_upregulated_GOterms <- top_upregulated_GOterms %>% filter(ONTOLOGY == "BP")
unique(top_upregulated_GOterms$GO)
```

```{r GO db}
# Load the package
library(GO.db)

# Look up the descriptions for the GO terms
GOterms_descriptions <- select(GO.db, keys = unique(top_upregulated_GOterms$GO), 
                               columns = "DEFINITION", keytype = "GOID")

head(GOterms_descriptions$DEFINITION)
```

Exercise 4.4
The gene symbol for the gene we just analysed is ZBTB16. Try to find information about this gene on the internet. What does this gene do?
  This gene is a member of the Krueppel C2H2-type zinc-finger protein family and encodes a zinc finger transcription factor that contains nine Kruppel-type zinc finger domains at the carboxyl terminus. This protein is located in the nucleus, is involved in cell cycle progression, and interacts with a histone deacetylase.[provided by RefSeq, Jul 2008]
  
  The gene encodes a transcription factor, which represses transcription. This is in line with the results of our GO term analysis.


Exercise 4.5
Use R to find the GO terms that are associated with DUSP1 (Entrez identifier: 1843). What is the function of DUSP1? Are any of the GO terms related to glucocorticoids (such as dexamethasone)?
```{r associated with DUSP1}

# Find the associated GO terms
dusp1_GOterms <- select(org.Hs.eg.db,
                        keys = "1843",
                        column = c("GO", "ONTOLOGY"),
                        keytype = "ENTREZID",
                        multiVals = "list")

# Select only the GO terms that are associated with biological processes
dusp1_GOterms <- dusp1_GOterms %>% filter(ONTOLOGY == "BP")

# Look up the descriptions for the GO terms
GOterms_descriptions <- select(GO.db, keys = unique(dusp1_GOterms$GO), 
                               columns = "DEFINITION", keytype = "GOID")

# Remove NA values
GOterms_descriptions <- GOterms_descriptions[!is.na(GOterms_descriptions$DEFINITION),]
head(GOterms_descriptions$DEFINITION, 10)

GOterms_descriptions$DEFINITION[str_detect(GOterms_descriptions$DEFINITION, "cortico")]
```

```{r GOstats}
# Load the GOstats library
library(GOstats)

# Create a list of upregulated genes
upregulated_genes <- airway_dge_results %>% data.frame() %>% 
  filter(log2FoldChange > 1, padj < 0.01) %>% rownames()

# Create a list of all genes in the dataset
all_genes <- airway_dge_results %>% data.frame() %>% rownames()

# Perform GO term enrichment analysis
test_object <- new("GOHyperGParams",
                   geneIds = upregulated_genes,
                   universeGeneIds = all_genes, 
                   annotation = "org.Hs.eg.db", 
                   ontology = "BP", 
                   pvalueCutoff = 1,
                   testDirection = "over")
goterm_analysis <- hyperGTest(test_object)

goterm_analysis
```

```{r }
# Obtains dataframe with results of GO term analysis
goterm_analysis_results <- summary(goterm_analysis)

# Adjust the p values for multiple testing
goterm_analysis_results$padj <- p.adjust(goterm_analysis_results$Pvalue, method = "BH")

# Select only gene sets that are larger than 5 but smaller 
# than 500 (to prevent taking along very small and very large
# gene sets).
goterm_analysis_results <- goterm_analysis_results %>% filter(Count > 5) %>% filter(Count < 500)

# Select the top 20 GO terms
goterm_analysis_top20 <- goterm_analysis_results[order(goterm_analysis_results$padj)[1:20],]

# Plot the p-values of the top 20 GO terms
goterm_analysis_top20$Term <- factor(goterm_analysis_top20$Term, 
                                     levels = goterm_analysis_top20$Term[
                                       order(goterm_analysis_top20$padj, decreasing = TRUE)])
goterm_analysis_top20 %>% ggplot(aes(x = Term, y = -log10(padj))) +
  geom_point() +
  coord_flip() +
  ylab(expression(-log[10](adjusted~italic(P)~value))) + 
  xlab("GO terms") +
  ggtitle("Top 20 enriched GO terms\n for upregulated genes") +
  theme_bw()
```

Exercise 4.6
Perform the GO term enrichment analysis for the downregulated genes. Use the genes with an adjusted p-value < 0.01 and a LFC < -1. Create a plot of the top 20 GO terms. What biological processes are associated with the downregulated genes?

```{r GO term enrichment analysis}

# Create a list of upregulated genes
downregulated_genes <- airway_dge_results %>% data.frame() %>% 
  filter(log2FoldChange < -1, padj < 0.01) %>% rownames()

# Create a list of all genes in the dataset
all_genes <- airway_dge_results %>% data.frame() %>% rownames()

# Perform GO term enrichment analysis
test_object2 <- new("GOHyperGParams",
                   geneIds = downregulated_genes,
                   universeGeneIds = all_genes, 
                   annotation = "org.Hs.eg.db", 
                   ontology = "BP", 
                   pvalueCutoff = 1,
                   testDirection = "over")
goterm_analysis2 <- hyperGTest(test_object2)

# Obtains dataframe with results of GO term analysis
goterm_analysis_results2 <- summary(goterm_analysis2)

# Adjust the p values for multiple testing
goterm_analysis_results2$padj <- p.adjust(goterm_analysis_results2$Pvalue, method = "BH")

# Select only gene sets with 5 < gene count < 500
goterm_analysis_results2 <- goterm_analysis_results2 %>% filter(Count > 5) %>% filter(Count < 500)

# Select the top 20 GO terms
goterm_analysis_top20 <- goterm_analysis_results2[order(goterm_analysis_results2$padj)[1:20],]

# Plot the p-values of the top 20 GO terms
goterm_analysis_top20$Term <- factor(goterm_analysis_top20$Term, 
                                     levels = goterm_analysis_top20$Term[
                                       order(goterm_analysis_top20$padj, decreasing = TRUE)])
goterm_analysis_top20 %>% ggplot(aes(x = Term, y = -log10(padj))) +
  geom_point() +
  coord_flip() +
  ylab(expression(-log[10](adjusted~italic(P)~value))) + 
  xlab("GO terms") +
  ggtitle("Top 20 enriched GO terms\n for downregulated genes") +
  theme_bw()

```